---
title: "Homework 6, Biostatistical Methods"
author: "Emil Hafeez (eh2928)"
date: "11/29/2020"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)

theme_set(theme_minimal() + theme(legend.position = "bottom")) #setup and establish the colors schemes for plits
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

First, let's read in the data.
```{r}
data_hosp = 
  read.csv("./data/PatSatisfaction.csv")
```

# Problem 1 (15p)

## Problem 1.1.
The correlation matrix refers to the array of numbers where \(r_{jk}\) is the pearson correlation coefficient between variables \(x_j\) and \(x_j\) such that 

$$
\mathbf{R}=\left(\begin{array}{ccccc}
1 & r_{12} & r_{13} & \cdots & r_{1 p} \\
r_{21} & 1 & r_{23} & \cdots & r_{2 p} \\
r_{31} & r_{32} & 1 & \cdots & r_{3 p} \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
r_{p 1} & r_{p 2} & r_{p 3} & \cdots & 1
\end{array}\right)
$$

As such, the correlation coefficient for the all variables "patient's satisfaction score" (the outcome), "age", "severity of illness", and "anxiety level" is as follows. 

```{r}
# Correlation matrix for all variables
cor(data_hosp) %>% 
  knitr::kable(
              align = "cccc", 
              digits = 3)
```



## Problem 1.2.

## Problem 1.3.

## Problem 1.4.

## Problem 1.5.









































# Problem 2 (15p)

## Problem 2.1.
## Problem 2.1.a.

## Problem 2.1.b.

## Problem 2.2.

## Problem 2.3.
## Problem 2.3.a.

## Problem 2.3.b.


